### 2024-06-14 manuscript.Rmd
revisions

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

Computational results, including those from statistical data analyses, are considered _reproducible_ if the same computation on the same data yields the same results if performed on a different computer or on the same computer later in time. Reproducibility is a prerequisite for replicable, robust and transparent research in digital environments.
 Various approaches have been suggested to increase chances of reproducibility [@peikert2021reproducible, @renv2022, @chan2023rang, @nagraj2023pracpac].
 Many approaches rely on R Markdown.
R Markdown is a simple language, which allows mixing natural language, simple formatting instructions (e.g., what is a headline, what should appear in bold face, what are list elements) and computer code (R, python or others) in a single document [@xie2018r].
When ever the document is compiled (or, _knitted_ in R speak), the computer code chunks are executed and allow for the dynamic generation of parts of the document.
A particular advantage of R Markdown is that it helps avoiding common threats to reproducibility of code, first and foremost, copy&paste errors between results of a regular R script and a scientific report (e.g., a Word document) [@peikert2021reproducible];
further, it is suitable to generate a variety of scientific assets such as posters [@posterdown2019], presentations [@xaringan2022], resumés [@vitae2023] or preregistrations [@peikert2021reproducible] from the same Markdown language.

The `reproducibleRchunk` package allows users to make computational results in R Markdown documents automatically testable for reproduction (does the same script with the same data produce the same results, e.g. on a different computer and/or later in time) with minimal changes to your workflow.
There is only a single thing users need to change in their Markdown document if they are already using R Markdown: In the first code chunk of their document, load the package and then change the code chunk type from `r` to `reproducibleR` for all code chunks that should be automatically checked for reproducibility.
In `reproducibleR` chunks, all newly declared variables are automatically identified and their contents are stored in a meta data file. 
Furthermore, the code chunk itself (that is, the code syntax) is also fingerprinted and stored, such that later non-reproduction can be traced back to either changes of the code chunk or changes of the computation.
Once a document is reproduced, in R-Markdown-speak _knitted_ again [@yihui2015], all computational results generated in the `reproducibleR` chunks are automatically tested for reproducibility.
The result of the reproduction attempt is displayed in a reproducibility report for each chunk.
Customizations of these reports are possible as we will demonstrate later.
A simple way to assess reproducibility of computational results would be  a bitwise comparison of the entire knitted document (i.e., the resulting HTML, PDF or DOCX file).
Yet, a particular advantage of using `reproducibleR` chunks is that users have a fine-grained control over which stages of a data analysis or any other computation should be checked for reproducibility. 
For example, if there is non-reproducibility, is it due to changes in the data file, during data cleaning or statistical modeling. 
Using `reproducibleRchunks`, users can separately assess the reproducibility of each stage of a given data analysis (or any other computation).
Also, not all elements of a research asset are required to reproduce (such as the current date displayed in a presentation of results).
In the following, we will briefly demonstrate how successful and failed reproduction will be checked and displayed by the package, we will go into details of the mechanics of the package, and discuss potential cases of non-reproducibility and how the package will handle these.

@@

The core functionality of the package is to provide a `reproducibleR` chunk, which are code chunks in R Markdown documents that can be used like regular `R`-chunks (including name labels and the usual options regulating output parameters). 
The package should be loaded in the first regular `R` code chunk in the document using
```{r eval=FALSE, echo=TRUE}
library(reproducibleRchunks)
```
On load, the package registers a new type of code chunk called `reproducibleR`.
The following example shows a snippet from an R Markdown file, in which there is a `reproducibleR` chunk with the name _addition_. 
In the chunk, a new variable `my_sum` is declared and defined to be the sum of a variable `x` plus one. 
Variable `x` was declared in an earlier code chunk (not shown in the following screenshot) and is thus not subject to the reproducibility test of this chunk.

```{r rstudio-screen1, fig.cap=""}
knitr::include_graphics("img/rstudio-screenshot-marker2.png")
```

In this example, the R Markdown document is knitted into a HTML-report.
On document generation, the R code in the `reproducibleR`-chunk is executed and the value of `my_sum` is stored in a separate file. 
A reproducibility report is shown below the code chunk to indicate all variables of the given chunk whose data are stored. The following is a snippet from the HTML generated on the first ever creation of the document:

```{r rstudio-screen2, fig.cap="Snippet from a knitted Markdown file showing the reproducibility report when a document is generated for the first time."}
knitr::include_graphics("img/generation-step1.png")
```
Once the document is generated a second time and the reproducibility data exists, the computation is executed anew and its results are compared to the stored reproducibility information on a variable by variable basis.
The reproducibility report will now contain information about the success of the individual reproduction attempts.
In our example, the computation was reproduced successfully:

```{r rstudio-screen3, fig.cap="Snippet from a knitted Markdown file showing the reproducibility report when a document is re-generated and a variable is successfully reproduced."}
knitr::include_graphics("img/generation-step2.png")

```
If the document is regenerated and the computational result is not identical to the original result, this failure of reproduction is indicated in the reproducibility report.
In this example, if someone changed the value of `x`, variable `my_sum` would change and the following failure would be displayed:

```{r rstudio-screen4, fig.cap="Snippet from a knitted Markdown file showing the reproducibility report when a document is re-generated and the reproduction of a variable failed."}
knitr::include_graphics("img/generation-step2-failed.png")

@@

For privacy reasons and to save disk space, we actually do not store the raw data by default but only fingerprints of the data.
Fingerprints are realized via so-called one-way hash functions, which can take an arbitrary large digital object and map it onto a fixed-size object, which is typically displayed as hexadecimal string.
Fingerprints allow for an efficient comparison of identity of results while avoid data security and privacy issues as they do not allow to reproduce the original data.
The following shows how the information about variable `numbers` is stored using a SHA256-fingerprint:

@@

```{r jsonschema, fig.cap="(ref:jsonschema) A JSON file stores fingerprints of all newly declared variables in a reproducibleR-chunk as well as fingerprints of the code chunk syntax."}
knitr::include_graphics("img/schema-json-fingerprints.png")

@@

In addition to the contents of all newly declared variables in a `reproducibleR`-chunk, the package also stores fingerprints of the code syntax.
This is shown in the schematic in Figure \@ref(jsonschema).

@@

What types of objects are suited for automated reproducibility tests? The answer is simple. In principle, any R object is suitable no matter how complex.
For creating fingerprints, the package uses R's serialization methods, which take any R object as input.
That is, all of the following variables (`x` of class _integer_, `y` of class _character_, and `z` of class _lm_) are valid examples for the automated reproducibility checks:

```{r examples, eval=FALSE, echo=TRUE}
  x <- 1:10
  y <- "qr"
  z <- lm(x~1, method=y)
```

@@

In the following, we provide a practical guide on how to use reproducible R code chunks in Markdown documents.

First, we define a reproducible code chunk by setting the code chunk language to `reproducibleR`. 
Once this document is rendered the first time, a data file will be created, which stores all reproducible results, which are computed in this code block. 
The name of the file will contain the label name (in this example `abc`). 
The following code block contains two reproducible results stored in variables `x` and `y`. 
Those computations can be based on results computed in previous chunks (here, variable `e`).
If the R markdown document is rendered a second time, the computational results are recomputed and compared against the original results. 
For each result, success or failure or reproduction is reported separately.
```{r eval=TRUE,echo=FALSE}
e<-1
```

```{reproducibleR abc,echo=TRUE,eval=TRUE}
xu <- rnorm(10, mean=0, sd=1) + e
y <- 4 * 4
```

If nothing goes wrong, rendering this markdown twice should result in a message of successful replication. 
To break reproducibility, remove or comment out the `set.seed()` command and render the document again. 
Now, you should obtain an error message indicating that the result of `x` could not be reproduced while `y` still reproduces as it is not dependend on the random number generator.

In practice, we expect cases, in which the reproduction report should be displayed (e.g., when testing and reproducing) or not (e.g., when rendering a presentation or a manuscript for submission to a journal). 
By default, reproduction report statements are produced for each chunk. 
Displaying these statements can be suppressed on a chunk-by-chunk basis by setting the chunk option

@@

Some default behavior of the package can be changed via R `options()`. The package defines the following options:

- reproducibleRchunks.digits: This is the number of digits for rounding of numbers and controls the numeric precision of the reproducibility checks. By default, this is $8$.
- reproducibleRchunks.filetype: By default this is 'json' 
- reproducibleRchunks.hashing:
- reproducibleRchunks.hashing_algorithm: This is the hashing algorithm, which is used to generate fingerprints of variable contents. By default, the `digest` package is used
- reproducibleRchunks.templates:

Here are a few examples how these options can be changed.

First, it is possible to change the precision with which numeric results are stored. By default, this is up to 8 digits. The following line reduces the precision to only four digits after the decimal point:

@@

If users wish to entirely suppress the default reproducibility reports, they can use the option `report=FALSE` as part of the code chunk options to suppress the display. 
Note that reproduction is still attempted and reproducibility information is still accessible through the function `get_reproducibility_summary()`.
The other standard chunk options can be used with `reproducibleR`-code chunks as usual. 
In particular, `eval=FALSE` suppressed execution of the code, `echo=FALSE` suppressed that the code is shown and so forth.

@@

- Original computations were executed and stored in the JSON data file. Later, a reproduction attempt is made using the exact same R Markdown file but on a different computer. This is a classic case of non-reproducibility, which often is due to changes in the software packages and the R version the computations in the R Markdown rely upon [@epskamp2019reproducibility, @peikert2021reproducible]. Note that the goal of this package is not to guarantee reproducibility but to allow for automated testing and reporting of reproducibility. To increase the chances of reproducibility in the first place, various solutions exist [@peikert2021reproducible, @renv2022, @chan2023rang, @nagraj2023pracpac] of which a primary aspect is a way to document and recreate the original computational environment.

@@

@@

Again, we emphasize that this package is not meant to ensure reproducibility but allows for automatically testing reproducibility.

Even though, a variety of approaches have been suggested to ensure reproducibility of computations in R Markdown documents [@peikert2021reproducible, @renv2022, @chan2023rang, @nagraj2023pracpac], we ourselves have encountered various situations, in which such approaches (including our own) failed. 
First and foremost, many approaches rely on further software packages such as Docker, which essentially provide virtual environments that run identical on different machines (or the same machine at different time points, e.g., before and after upgrade of R or some or all packages used).
On some machines, Docker may simply be not available, either because it is not (yet) available on a certain newly introduced hardware (e.g., when Apple switched to their own processors) or because a user does not have admin privileges on a machine.
Further some approaches rely on service providers such as the MRAN archive, which was unexpectedly terminated. 
These different Achilles' heels at least occasionally lead to situations, in which users will try to locally reproduce historic computational results.
With our package, the success of these reproductions becomes easily testable.

>>>>

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed,
                      out.width = "80%")
```

Computational results, including those from statistical data analyses, are considered _reproducible_ if the same computation on the same data yields the same results if performed on a different computer or on the same computer later in time. Reproducibility is a prerequisite for replicable, robust and transparent research in digital environments [@epskamp2019reproducibility].
 Various approaches have been suggested to increase chances of reproducibility [@peikert2021reproducible; @renv2022; @chan2023rang; @nagraj2023pracpac].
 Many approaches rely on R Markdown [@rmarkdown2023].
R Markdown is a simple language, which allows mixing natural language, simple formatting instructions (e.g., what is a headline, what should appear in bold face, what are list elements) and computer code (R, python or others) in a single document [@xie2018r].
When ever the document is compiled (or, _knitted_ in R speak), all of its computer code chunks are executed and thus generate dynamic content in the document.
This is particularly interesting for psychological research reports, as the `papaja` package [@R-papaja] allows R Markdown documents (including tables and figures) to be automatically formatted in APA style; most importantly, statistical results and content of tables and figures are dynamically generated.
Also, (R) Markdown forms the basis of _executable research articles_ [@tsang2020welcome], a browser-based form of research articles that allows authors to enrich their publications with interactive elements, such as figures or results that dynamically change based on readers' input.
A particular advantage of using R Markdown for creating research reports is that it helps avoiding common threats to reproducibility of statistical results, first and foremost, copy&paste errors between results of a regular R script and a scientific report (e.g., a statistic or table in a Word document) [@peikert2021reproducible];
further, it is suitable to generate a variety of scientific assets other than manuscripts such as posters [@posterdown2019], presentations [@xaringan2022], resumés [@vitae2023] or preregistrations [@peikert2021reproducible] from the same Markdown language.

The `reproducibleRchunk` package allows users to make computational results in R Markdown documents automatically testable for reproduction (does the same script with the same data produce the same results, e.g. on a different computer and/or later in time) with minimal changes to your workflow.
There is only a single thing users need to change in their Markdown document if they are already using R Markdown: In the first code chunk of their document, load the package and then change the code chunk type from `r` to `reproducibleR` for all code chunks that should be automatically checked for reproducibility.
In `reproducibleR` chunks, all newly declared variables are automatically identified and their contents are stored in a meta data file. 
Furthermore, the code chunk itself (that is, the code syntax) is also fingerprinted and stored, such that later non-reproduction can be traced back to either changes of the code chunk or changes of the computation.
Once a document is reproduced, in R-Markdown-speak _knitted_ again [@yihui2015], all computational results generated in the `reproducibleR` chunks are automatically tested for reproducibility.
The result of the reproduction attempt is displayed in a reproducibility report for each chunk.
Customizations of these reports are possible as we will demonstrate later.
A simple way to assess reproducibility of computational results would be  a bitwise comparison of the entire knitted document (i.e., the resulting HTML, PDF or DOCX file).
Yet, a particular advantage of using `reproducibleR` chunks is that users have a fine-grained control over which stages of a data analysis or any other computation should be checked for reproducibility. 
For example, if there is non-reproducibility, is it due to changes in the data file, during data cleaning or statistical modeling. 
Using `reproducibleRchunks`, users can separately assess the reproducibility of each stage of a given statistical data analysis (or any other computation).
Also, not all elements of a research asset are required to reproduce (such as the current date displayed in a presentation of results).
In the following, we will briefly demonstrate how successful and failed reproduction will be checked and displayed by the package, we will go into details of the mechanics of the package, and discuss potential cases of non-reproducibility and how the package will support researchers in handling these cases.

@@

The core functionality of the package is to provide a `reproducibleR` chunk, which are code chunks in R Markdown documents that can be used like regular `R`-chunks (including name labels and the usual options regulating output parameters). 
To use these new code chunks, the package must be loaded in the first regular `R` code chunk in the document using

```{r eval=FALSE, echo=TRUE}
library(reproducibleRchunks)
```

On load, the package registers a new type of code chunk called `reproducibleR`.
Figure \@ref(fig:rstudioscreen1) shows a snippet from an R Markdown file, in which there is a `reproducibleR` chunk with the name _addition_. 
In the chunk, a new variable `my_sum` is declared and defined to be the sum of some variable `x` plus one. 
Variable `x` was declared in an earlier code chunk (not shown in the figure) and is thus not subject to the reproducibility test of this chunk.

```{r rstudioscreen1, fig.cap="Reproducibility report shown when a document is created for the first time and no reproducibility information is available yet."}
knitr::include_graphics("img/rstudio-screenshot-marker2.png")
```

```{r rstudioscreen2, fig.cap="Snippet from a knitted Markdown file showing the reproducibility report when a document is generated for the first time."}
knitr::include_graphics("img/generation-step1.png")
```

In this example, we assume that the R Markdown document is eventually knitted into a HTML-report (but all other pandoc-supported formats are supported as well).
On document generation, the R code in the `reproducibleR`-chunk is executed and the value of `my_sum` is stored in a separate file, which contains all information for further reproducibility tests. 
A reproducibility report is shown in the generated document below the code chunk indicating all variables of the given chunk whose data are stored. Figure \@ref(fig:rstudioscreen2) show a snippet from the HTML report generated on the first ever creation of the document.

Once the document is generated a second time and the reproducibility data exists, the computation is executed anew and its results are compared to the stored reproducibility information on a variable by variable basis.
The reproducibility report will now contain information about the success of the individual reproduction attempts.
In our example, the computation was reproduced successfully.
A snippet from the resulting generated document is shown in Figure \@ref(fig:rstudioscreen3).

```{r rstudioscreen3, fig.cap="Snippet from a knitted Markdown file showing the reproducibility report when a document is re-generated and a variable is successfully reproduced."}
knitr::include_graphics("img/generation-step2.png")

```
If the document is regenerated and the computational result is not identical to the original result, this failure of reproduction is indicated in the reproducibility report.
In a variation of the previous example, we assume someone changed the value of `x`, then variable `my_sum` would change and the a failure would be displayed (see Figure \@ref(fig:rstudioscreen4)).

```{r rstudioscreen4, fig.cap="Snippet from a knitted Markdown file showing the reproducibility report when a document is re-generated and the reproduction of a variable failed."}
knitr::include_graphics("img/generation-step2-failed.png")

@@

For privacy reasons and to save disk space, we actually do not store the raw data by default but only fingerprints of the data.
Fingerprints are realized via so-called one-way hash functions, which can take an arbitrary large digital object and map it onto a fixed-size object, which is typically displayed as hexadecimal string.
Fingerprints allow for an efficient comparison of identity of results while avoid data security and privacy issues as they do not allow to reproduce the original data.
The following shows how the information about variable `numbers` is stored using a SHA256-fingerprint (which is a 256 bit-long fingerprint that is usually displayed using 64 hexadecimal characters):

@@

```{r jsonschema, fig.cap="A JSON file stores fingerprints of all newly declared variables in a reproducibleR-chunk as well as fingerprints of the code chunk syntax. This image by Andreas Brandmaier and Anke Schneider is licensed under CC BY 4.0"}
knitr::include_graphics("img/schema-json-fingerprints_v4.png")

@@

In addition to the contents of all newly declared variables in a `reproducibleR`-chunk, the package also stores a fingerprint of the code syntax.
This allows later checks whether a potential non-reproduction is due to a modification of the code chunk since the original computation.
The storage of the fingerprinted information about a code chunk and its computational results is illustrated in the schematic in Figure \@ref(fig:jsonschema).

@@

What types of objects are suited for automated reproducibility tests? The answer is simple. In principle, any R object is suitable no matter how complex.
For creating fingerprints, the package uses R's serialization methods, which take any R object as input.
That is, all of the following variables (`x` of class _integer_, `y` of class _character_, `z` of class _lm_, and `lst` of class _list_) are valid examples for the automated reproducibility checks:

```{r examples, eval=FALSE, echo=TRUE}
  x <- 1:10
  y <- "qr"
  z <- lm(x~1, method=y)
  lst <- list(x, y, z)
```

Yet, we recommend subjecting simple objects to reproducibility checks, that is, preferably numeric variables such as single numbers or vectors of numbers. 
In a typical statistical data analysis this would correspond to, for example, parameter estimates, effect size estimates, confidence interval limits, p values, fit indices, or Bayes factors.
While virtually any R object is serializable in principle, we expect that the serialization algorithm is more likely to change over time for more complex objects in future than for simple objects.

@@

In the following, we provide a practical guide on how to use reproducible R code chunks in Markdown documents. First, to install the package, please install the latest version from our repository:

```{r eval=FALSE, echo=TRUE}
devtools::install_github("brandmaier/reproducibleRchunks")
```

In your R Markdown file, load the package using  `library(reproducibleRchunks)`, preferably in the first regular R code chunk at the very top of the document.

Now, we define a reproducible code chunk by setting the code chunk language to `reproducibleR`. 
Once this document is rendered the first time, a data file will be created, which stores all reproducible results, which are computed in this code block. 
The name of the file will contain the label name (in this example `helloworld`). 
The following code block contains two reproducible results stored in variables `x` and `y`. 
Those computations can be based on results computed in previous chunks but those variables will not per se be subject to the reproducibility tests.
If the R markdown document is rendered a second time, the computational results are recomputed and compared against the original results. 
For each result, success or failure or reproduction is reported separately.

<!--
```{reproducibleR helloworld, echo=TRUE, eval=TRUE}
set.seed(42)
x <- rnorm(10, mean=0, sd=1) 
y <- 4 * 4
```
-->
```{r fig.cap="A reproducibleR chunk with two variables automatically tested for reproducibility."}
knitr::include_graphics("img/helloworld.png")
```

Rendering this markdown twice in a row on the same computer should result in a message of successful replication. 
To break reproducibility, remove or comment out the `set.seed()` command, which sets the random number generator in a reproducible state, and render the document again. 
Now, you should obtain an error message indicating that the result of `x` could not be reproduced while `y` still reproduces as it did not depend on the random number generator.

In practice, we expect cases, in which the reproduction report should be displayed (e.g., when developing a data analysis script or reproducing a former one) or not (e.g., when rendering a presentation or a manuscript for submission to a journal). 
By default, reproduction report statements are produced for each chunk but this default can be changed via the code chunk argument `report=FALSE`.
If users wish to change the default display of reports over the entire document, this can be adjusted via a global knitr option as follows:

```{r echo=TRUE, eval=FALSE}
knitr::opts_chunk$set(report = FALSE)
```

@@

Some default behaviors of the package can be changed via R `options()`. The package defines the following options:

- reproducibleRchunks.digits: This is the number of digits for rounding of numbers and controls the numeric precision of the reproducibility checks. By default, this is $8$.
- reproducibleRchunks.filetype: This is the type of data storage. Currently, we only support the 'json' format; however, this leaves room for other formats to be supported in future. 
- reproducibleRchunks.hashing: Boolean. Indicates whether fingerprints should be used (default) or raw values should be stored.
- reproducibleRchunks.hashing_algorithm: This is the hashing algorithm, which is used to generate fingerprints of variable contents. By default, the algorithm offered by the `digest` package are used.
- reproducibleRchunks.templates: A list with keys corresponding to pandoc output formats and values to templates for formatting the reproducibility reports. See below for more details.

Here are a few examples how these options can be changed.
First, it is possible to change the precision with which numeric results are stored. By default, this is up to 8 digits. The following line reduces the precision to only four digits after the decimal point:

@@

As mentioned above, if users wish to entirely suppress the default reproducibility reports, they can use the chunk argument `report=FALSE` as part of the code chunk options to suppress the display. 
Note that reproduction is still attempted and reproducibility information is still accessible through the function `get_reproducibility_summary()`.
The other standard chunk options can be used with `reproducibleR`-code chunks as usual. 
In particular, `eval=FALSE` suppressed execution of the code, `echo=FALSE` suppressed that the code is shown and so forth.

@@

- Original computations were executed and stored in the JSON data file. Later, a reproduction attempt is made using the exact same R Markdown file but on a different computer. This is a classic case of non-reproducibility, which often is due to changes in the software packages and the R version the computations in the R Markdown rely upon [@epskamp2019reproducibility; @peikert2021reproducible]. Note that the goal of this package is not to guarantee reproducibility but to allow for automated testing and reporting of reproducibility. To increase the chances of reproducibility in the first place, various solutions exist [@peikert2021reproducible; @renv2022; @chan2023rang; @nagraj2023pracpac] of which a primary aspect is a way to document and recreate the original computational environment.

@@

- A future `R` version changes the way objects are serialized (that is, converted from the internal representation to a byte-stream representation, of which the fingerprint is taken). The version of the serialize protocol can be manually overwritten to an older version if newer versions should not be backward compatible. As long as the `digest` package is used for generating fingerprints, the serialization version can be fixed to the version that is used at the time of writing (2024) using this command: `options(serializeVersion=2)`.

<!-- - The results of hash functions are generally not guaranteed to be stable over time. To trace back possible problems, -->

@@

Again, we emphasize that this package is not meant to ensure reproducibility but allows for automatically testing reproducibility.
Even though, a variety of approaches have been suggested to ensure reproducibility of computations in R Markdown documents [@peikert2021reproducible; @renv2022; @chan2023rang; @nagraj2023pracpac], we ourselves have encountered various situations, in which such approaches (including our own) failed. 
First and foremost, many approaches rely on further software packages such as Docker, which essentially provide virtual environments that run identical on different machines (or the same machine at different time points, e.g., before and after upgrade of R or some or all packages used).
On some machines, Docker may simply be not available, either because it is not (yet) available on a certain newly introduced hardware (e.g., when Apple switched to their own processors) or because a user does not have admin privileges on a machine.
Further some approaches rely on service providers such as the MRAN archive, which was unexpectedly terminated. 
These different Achilles' heels at least occasionally lead to situations, in which users will try to locally reproduce historic computational results.
With our package, the success of these reproductions becomes easily testable.
